Instantiating Bunched Dependent Type Theory for Monoidal Classifying Toposes

### 2025-06-30 paper.tex
Clean up proof. Completes handling of issue #2.

% \usepackage[style=alphabetic]{biblatex}
% \addbibresource{paper.bib}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathpartir}

@@

\begin{abstract}
    In their PhD thesis, Sch\"opp introduces a type theory designed to capture the monoidal structure in the Schanuel topos.
    I provide a simple condition that can be used to verify whether the type theory interprets into a classifying topos of your choice.
\end{abstract}

@@

I provide an easy method to verify that \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\)
can be interpreted in a classifying topos of your choice,
so you can start \emph{using} the internal language as quickly as possible.

@@

For simplicity of exposition, I describe the theorem and proof in the case of single-sorted theories.
The extension to multiple sorts is straightforward, but terminologically awkward.

\begin{theorem}
    Let \(\T\) be a single-sorted geometric theory, with sort \(X\).
    Let \((\sim) : X \to X \to \Prop\) be a geometrically-defined symmetric relation.
    Further suppose that for any \(x \in X\), \(\{a : X \mid a \sim x\}\) is a sub-\(\T\)-model of \(X\).
    Then \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\) interprets into \(\Set[\T]\).
\end{theorem}

For example, let \(\T\) be the theory of rings, and let \(a \sim b\) mean \(ab = ba\).
The hypothesis holds because centralizers of rings are rings;
so we get an interpretation of the bunched type theory into \(\Set[\Ring]\).

@@

As an example of the latter, say \(\T\) is the theory of inhabited sets, with axiom \(\vdash \exists a : X, \top\).
Then the required sequent is \(x : X \vdash \exists a : X, a \sim x \land \top\).

@@

For the duration of this section, we accept the assumptions of the main theorem.

\noindent To begin, notice a key detail of the transformation described above.
\begin{lemma} \label{transform-lemma}
    The transformation of sequents described in section \ref{Details},
    used there to transform the axioms of \(\T\), preserves derivability.
    Any provable sequent remains provable after the transformation.
\end{lemma}

@@

Next, bootstrap to the syntactic site.
\begin{construction}
    We place a monoidal structure on \(\T\)'s syntactic category.
\end{construction}

@@

    We've defined \(f * 1_B\). The definition of \(1_A * g\) is analogous.
    All of the functoriality laws, including the interchange condition \((f * 1) \circ (1 * g) = (1 * g) \circ (f * 1)\),
    follow from relating \(f * 1_B\) back to \(f \times 1_B\), and using the bifunctoriality of the cartesian product.
\end{proof}

@@

Lift to the classifying topos.

\begin{construction} \label{sh}
    We equip \(\Set[\T]\) with a strictly affine symmetric monoidal closed structure \((*,\wand)\).
\end{construction}

@@

    Expanding both \(\otimes^\Psh\) and \(\times\) as Day convolutions,
    this is a \(\Set\)-map, natural in \(c \in C^\mathrm{op}\), of the following type.
    \begin{scriptsize}\[\left(\int^{c_1,c_2 \in C} A(c_1) \times B(c_2) \times (c \xrightarrow{C} c_1 * c_2)\right) \to \left(\int^{c_1,c_2 \in C} A(c_1) \times B(c_2) \times (c \xrightarrow{C} c_1 \times c_2)\right)\]\end{scriptsize}
    This map is exactly what you'd expect it to be; it composes the \(c \xrightarrow{C} c_1 * c_2\) component
    with the canonical map \(c_1 * c_2 \xrightarrow{C} c_1 \times c_2\), and leaves everything else alone.
    Since that canonical map is mono, this is too.
\end{proof}

And finally:

\begin{construction}
    We have an interpretation of \(\Set[\T]\) into \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\).
\end{construction}

\begin{proof}
    Apply section 3.5 of \cite{schopp} to see that \(\Set[\T]\) is a ``\((*, 1, \Sigma, \Pi, \Pi^*)\)-type-category'',
    then apply Chapter 6 of the same to construct the interpretation.
\end{proof}

\section{Using the Monoidal Structure}

For practical use, the main theorem isn't quite enough.
You'll need to know not just that \(*\)-types exist, but how they work.
If \(R\) is the generic local ring, you'll want to know that \(R * R = \{(a,b) : R^2 \mid ab = ba\}\).

Hence this next theorem.

\begin{theorem}
    Let \(U\) be the generic \(\T\)-model in \(\Set[\T]\).
    Then \(U * U = \{(a,b) : U \times U \mid a \sim b\}\), as subsets of \(U \times U\).
\end{theorem}

\begin{proof}
    Viewing \(\Set[\T]\) as \(\Sh(\C_\T)\), expand.
    \begin{align*}
        U \otimes^\Sh U
        &\simeq \mathbf{a}\yo(x:X) \otimes^\Sh \mathbf{a}\yo(x:X)
        \\ &\simeq \mathbf{a}(i\mathbf{a}\yo(x:X) \otimes^\Psh i\mathbf{a}\yo(x:X))
        \\ &\simeq \mathbf{a}(\yo(x:X) \otimes^\Psh \yo(x:X))
        \\ &\simeq \mathbf{a}\yo((x:X) * (x:X))
        \\ &\simeq \mathbf{a}\yo(a:X, b:X, a \sim b)
        \\ &\simeq \{(a,b) : U \times U \mid a \sim b\}
    \end{align*}

@@

\section{Backwards}

TODO: Trying to make the main theorem an iff statement
TODO: This entire section will need to be rewritten once the proof works.
POSSIBILITY: Make the proof a chain of iffs. That way, there's a non-classifying characterization of topos models as well as a classifying one.
% culminating in a beautiful characterization in terms of classifying toposes.

Consider any model of \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\) in the classifying topos over \(\T\).

TODO: Translate properties of the type theory to properties of the topos. E.g. \(*\) affine monoidal product...

Let \(U\) be the generic \(\T\)-model in \(\Set[\T]\).

\begin{lemma}
    For any \(A \in \Set[\T]\), the data \(A * -, \Pi^*_A\) is determined by the subset \(A * U \subseteq A \times U\).
\end{lemma}

\begin{proof}
    Observe that TODO form a geometric morphism \(\Set[\T]_{/A} \to \Set[\T]\).
    Each such geometric morphism yields a \(\T\)-model in \(\Set[\T]_{/A}\),
    given by the restriction of the obvious model \(\substack{A \times U \\ \downarrow \\ A}\) to \(\substack{A * U \\ \downarrow \\ A}\).
\end{proof}

@@

    The former is immediate. The latter holds because the geometric morphism TODO induces a \(\T\)-model in \(\Set[\T]_{/A}\).
\end{proof}

@@

% \printbibliography

>>>>

\usepackage[style=alphabetic]{biblatex}
\addbibresource{paper.bib}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{quiver}

\usepackage{todonotes}

@@

\begin{abstract}
    In their PhD thesis, Sch\"opp introduces a type theory designed to capture the monoidal structure in the Schanuel topos.
    I provide a strikingly simple characterization of the models of this type theory in arbitrary classifying toposes.
\end{abstract}

\listoftodos

@@

I characterize the models of \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\) in an arbitrary classifying topos.
This characterization is extremely easy to use; describing a topos model typically becomes a single-paragraph task.

For simplicity, I describe only the case of single-sorted type theories.
While both the theorem and proof extend naturally to multi-sort theories,
the notation becomes more awkward in that case.

@@

Recall from \cite{schopp}, section 3.5 and chapter 6, that topos models of the theories
\(\mathbf{BT}(1, \Sigma, \Pi, \Pi^*)\), \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\), or \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*, B^{*(M:A)})\)
come from toposes equipped with \emph{strict affine symmetric monoidal closed structures that preserve pullback}.
That is, symmetric monoidal structures \((*,I)\) where:
\begin{itemize}
    \item \(I\) is terminal.
    \item The canonical map \(A * B \to A \times B\) is mono.
    \item The functor \(A * -\) preserves pullbacks, and has a right adjoint.
\end{itemize}

I characterize these structures in classifying toposes.

\begin{theorem}
    Let \(\T\) be a single-sorted geometric theory, with sort \(X\).
    Monoidal structures on topos \(\Set[\T]\) of the above form 
    are given by symmetric relations \((\sim) : X \to X \to \Prop\),
    satisfying the following property.
    \begin{quote}
        For any \(x : X\), the \(\T\)-model structure on \(X\)
        restricts to a sub-\(\T\)-model on \(\{a : X \mid a \sim x\}\).
    \end{quote}

    Under this interpretation, if \(U\) is the generic \(\T\)-model,
    we will have \(U*U = \{(a,b) : U \times U \mid a \sim b\}\).
\end{theorem}

\noindent For example, to interpret the type theory into \(\Set[\Ring]\), let \(a \sim b\) mean \(ab = ba\).
We need only check that \(\{a : R \mid ax = xa\}\) is a ring for any \(x\), and this holds by standard ring theory.
The monoidal product of the generic ring with itself will be the set of commuting pairs of ring elements.

% \begin{theorem}
%     Let \((\sim) : X \to X \to \Prop\) be a symmetric relation, geometrically defined in the language of the theory.
%     Further suppose that for any \(x \in X\), \(\{a : X \mid a \sim x\}\) is a sub-\(\T\)-model of \(X\).
%     Then \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\) interprets into \(\Set[\T]\).
% \end{theorem}

% For example, if \(\T\) is the theory of rings, let \(a \sim b\) mean \(ab = ba\).
% The hypothesis holds because centralizers of rings are rings,
% so we get an interpretation of the bunched type theory into \(\Set[\Ring]\).

@@

As an example of the latter condition, say \(\T\) is the theory of inhabited sets, with axiom \(\vdash \exists a : X, \top\).
Then the required sequent is \(x : X \vdash \exists a : X, a \sim x \land \top\).

@@

\subsection{Notation}

We use the following notation.
\begin{itemize}
    \item \(\T\) will be a single-sorted geometric theory, with sort \(X\).
    \item \(\C_\T\) will be the syntactic site of \(\T\).
    \item \(\Psh(\C_\T)\) and \(\Sh(\C_\T)\) are the categories of presheaves and sheaves on the syntactic site.
    The latter is the classifying topos of \(\T\), so we also call it \(\Set[\T]\).
    \item We have the following diagram of functors:
    % https://q.uiver.app/#q=WzAsMyxbMCwwLCJcXENfXFxUIl0sWzIsMCwiXFxQc2goXFxDX1xcVCkiXSxbNCwwLCJcXFNoKFxcQ19cXFQpID0gXFxTZXRbXFxUXSJdLFswLDEsIlxceW8iXSxbMSwyLCJcXG1hdGhiZnthfSIsMCx7Im9mZnNldCI6LTJ9XSxbMiwxLCJpIiwwLHsib2Zmc2V0IjotMn1dLFs0LDUsIiIsMCx7ImxldmVsIjoxLCJzdHlsZSI6eyJuYW1lIjoiYWRqdW5jdGlvbiJ9fV1d
    \[\begin{tikzcd}
        {\C_\T} && {\Psh(\C_\T)} && {\Sh(\C_\T) = \Set[\T]}
        \arrow["\yo", from=1-1, to=1-3]
        \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathbf{a}}", shift left=2, from=1-3, to=1-5]
        \arrow[""{name=1, anchor=center, inner sep=0}, "i", shift left=2, from=1-5, to=1-3]
        \arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \end{tikzcd}\]
    \(\yo\) is the Yoneda embedding, \(\mathbf{a}\) is sheafification, and \(i\) is the inclusion of presheaves into sheaves.
    The composite functor \(\mathbf{a} \circ \yo\) interprets the syntax of the theory into the generic model.
    \item \(U = \mathbf{a}\yo(x:X)\) is the interpretation of the \(\T\)'s sort in the generic model.
\end{itemize}

\subsection{Forwards Direction}

For the duration of this section, let \((\sim) : X \to X \to \Prop\) be a relation satisfying the property described in the main theorem.

\begin{lemma} \label{transform-lemma}
    The transformation of sequents described in section \ref{Details},
    used there to transform the axioms of \(\T\), preserves derivability.
    Any provable sequent remains provable after the transformation.
\end{lemma}

@@

\begin{construction}
    We place a monoidal structure on \(\T\)'s syntactic category.
\end{construction}

@@

    We've defined \(f * 1_B\). The definition of \(1_A * g\) is analogous.
    All of the functoriality laws, including the interchange condition \((f * 1) \circ (1 * g) = (1 * g) \circ (f * 1)\),
    follow from relating \(f * 1_B\) back to \(f \times 1_B\), and using the functoriality of the cartesian product.
\end{proof}

@@

\begin{construction} \label{sh}
    We equip \(\Set[\T]\) with a pullback-preserving strict affine symmetric monoidal closed structure.
\end{construction}

@@

    Expanding both \(\otimes^\Psh\) and \(\times\) as Day convolutions,
    this is a \(\Set\)-map, natural in \(c \in C^\mathrm{op}\), of the following type.
    \begin{scriptsize}\[\left(\int^{c_1,c_2 \in C} A(c_1) \times B(c_2) \times (c \xrightarrow{C} c_1 * c_2)\right) \to \left(\int^{c_1,c_2 \in C} A(c_1) \times B(c_2) \times (c \xrightarrow{C} c_1 \times c_2)\right)\]\end{scriptsize}
    This map is exactly what you'd expect it to be; it composes the \(c \xrightarrow{C} c_1 * c_2\) component
    with the canonical map \(c_1 * c_2 \xrightarrow{C} c_1 \times c_2\), and leaves everything else alone.
    Since that canonical map is mono, this is too.

    \todo{Missing condition: Pullback preservation. I'm not sure how to prove this, or if I need to change the theorem statement to make it work.}
\end{proof}

\begin{theorem}
    Let \(U\) be the generic \(\T\)-model in \(\Set[\T]\).
    Then \(U \otimes^\Sh U = \{(a,b) : U \times U \mid a \sim b\}\), as subsets of \(U \times U\).
\end{theorem}

\begin{proof}
    \begin{align*}
        U \otimes^\Sh U
        &\simeq \mathbf{a}\yo(x:X) \otimes^\Sh \mathbf{a}\yo(x:X)
        \\ &\simeq \mathbf{a}(i\mathbf{a}\yo(x:X) \otimes^\Psh i\mathbf{a}\yo(x:X))
        \\ &\simeq \mathbf{a}(\yo(x:X) \otimes^\Psh \yo(x:X))
        \\ &\simeq \mathbf{a}\yo((x:X) * (x:X))
        \\ &\simeq \mathbf{a}\yo(a:X, b:X, a \sim b)
        \\ &\simeq \{(a,b) : U \times U \mid a \sim b\}
    \end{align*}

@@

\subsection{Backwards Direction}

Consider any pullback-preserving strict affine symmetric monoidal structure on \(\Set[\T]\).

\begin{lemma}
    For any \(A \in \Set[\T]\), there is a geometric morphism \(W_A \dashv \Pi^*_A : \Set[\T]_{/A} \leftrightarrows \Set[\T]\),
    whose composition with \(\Sigma_A \dashv A^* : \Set[\T] \leftrightarrows \Set[\T]_{/A}\)
    yields the adjunction \((A * -) \dashv (A \wand -)\).
\end{lemma}

\begin{proof}
    Following the construction in section 2.5.1 of \cite{schopp}, we define:
    \begin{align*}
        W_A(B) &= \substack{A * B \\ \downarrow \\ A}
        \\ \Pi^*_A\left(\substack{B \\ \enspace \downarrow f \\ A}\right) &= \text{Pullback of } (A \wand B \xrightarrow{A \wand f} A \wand A \xleftarrow{id} 1)
    \end{align*}
\end{proof}

\begin{corollary}
    For any \(A \in \Set[\T]\), the functor \(A * -\) is determined by \(A * U\).
\end{corollary}

\begin{proof}
    The above geometric morphism corresponds to a \(\T\)-model in \(\Set[\T]_{/A}\),
    given by the restriction of the obvious model \(\substack{A \times U \\ \downarrow \\ A}\) to \(\substack{A * U \\ \downarrow \\ A}\).
\end{proof}

@@

    The former is immediate. The latter holds because the geometric morphism \(W_U \dashv \Pi^*_U\) induces a \(\T\)-model in \(\Set[\T]_{/U}\).
\end{proof}

@@

\printbibliography
